A fast algorithm for global code motion of congruent instructions

### 2017-11-13 cgo/paper.tex
call and fast algo

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
very well on single basic-blocks (local level) but when it comes to detect
redundancies across basic-blocks (global level) these techniques fall short:
more complex passes like Global Common Subexpression Elimination (GCSE) and
Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
based on data-flow analysis \cite{morel1979global}.  At first these techniques
were described in the classical data-flow analysis framework, but later the use
of Static Single Assignment (\SSA{}) representation lowered their cost in terms
of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
and brought these techniques in the main stream: \SSA{} based \PRE{} is
available in many compilers like \GCC{} and \LLVM{}.

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., calls with missing attributes.

\begin{lstlisting}
// Compute barriers for both hoistable and sinkable
// candidates.
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions
    // can be sunk. If there was no barrier in B, any
    // instruction satisfying other legality checks
    // can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{lstlisting}

Computing barriers allows for efficient removal of non-movable candidates, this
makes the legality check converge faster.

@@

We need to prove that on all execution paths from the initial position of an
instruction to its destination, the side effects appear in the same order. In
general, we cannot introduce a new computation along any path of execution
without checking for undefined behavior \cite{undef}. It is also necessary to
check if there are indirect branch targets, e.g., landing pad, case statements,
goto labels, etc., along the path because it becomes difficult to prove safety
checks in those cases.

@@

\subsubsection{Hoisting an expression across a call}
\label{cost:across-calls}
Even hoisting scalars across calls is tricky because it can increase the number
of spills. During the frame lowering of calls, the argument registers (also
known as the caller saved registers) are saved because they might be modified by
the callee and after the call they are restored. Due to this the register
pressure is high because the number of available registers are reduced by the
number of caller saved registers. In that situation a computation that increases
register pressure is not profitable to hoist.

\section{A fast algorithm for code motion of congruent computations}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for code motion. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed.

@@

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). If the dependency of a hoistable candidate is in the
same basic-block as the candidate, then the dependency must also be hoistable,
otherwise hoisting will be illegal or would require a complicated code
generation to make it legal. The current algorithm discards cases if the
dependency is neither hoistable nor rematerializable. We collect the \GVN{} of
all the instructions in the function and iterate on the list of instructions
having identical \GVN{}s. The algorithm \emph{doGCM} prefers hoisting
(\emph{doCodeHoisting}) to sinking (\emph{doCodeSinking}).

>>>>

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
very well on single basic-blocks (local level) but when it comes to detect
redundancies across basic-blocks (global level) these techniques fall short:
more complex passes like Global Common Subexpression Elimination (GCSE) and
Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
based on data-flow analysis \cite{morel1979global}.  At first these techniques
were described in the classical data-flow analysis framework, but later the use
of Static Single Assignment (\SSA{}) representation lowered their cost in terms
of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
and brought these techniques in the main stream: \SSA{} based \PRE{} has been
implemented in many optimizing compilers like \GCC{} and \LLVM{}.

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. A barrier will prevent any instruction occurring lexically
below it from getting hoisted. On the other hand, a sink barrier will prevent
any instruction above it from getting sunk. If there are no barriers in a basic
block, any instruction satisfying other legality checks can be sunk.

\begin{lstlisting}
// Compute both hoist and sink barriers
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

    // Find the last barrier below which instructions
    // can be sunk.
    if barrier_found:
      for each instruction I in B in reverse order:
        if I does not guarantee progress:
          mark I as a sink barrier
          break
\end{lstlisting}

In the absence of context, as in our current implementation, some instructions
which might be safe are still classified as barriers, e.g., calls with missing
attributes.  Computing barriers allows for efficient removal of non-movable
candidates, this makes the legality check converge faster.

@@

We need to prove that on all execution paths from the initial position of an
instruction to its destination, the side effects appear in the same order. In
general, we cannot introduce a new computation along any path of execution
without checking for undefined behavior \cite{undef}. It is also necessary to
check if there are indirect branch targets, e.g., landing pad, case statements,
or other forms of goto labels along the path because it becomes difficult to
prove safety checks in those cases.

@@

\subsubsection{Hoisting an expression across a call}
\label{cost:across-calls}
Call expressions are difficult to analyze in an intraprocedural optimization
pass like \GCM{}. Even hoisting scalars across calls is tricky because it can
increase the number of spills. During the register allocation of calls
expressions, the argument registers (also known as the caller saved registers)
are saved because they might be modified by the callee and after the call they
are restored. Due to this the register pressure is high because the number of
available registers are reduced by the number of caller saved registers. In that
situation a computation that reduces register pressure may be beneficial to
hoist. Similarly, a computation that introduces a register (without killing any
other operand) before a call may be beneficial to sink.

\section{A fast algorithm for code motion of congruent computations}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for code motion. A copy of
the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed and affected data structuers are
updated.

@@

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions which makes it easier to ensure
that we do not introduce a redundant computation along any path of execution. If
the dependency of a hoistable candidate is in the same basic-block as the
candidate, then the dependency must also be hoistable, otherwise hoisting will
be illegal or would require a complicated code generation to make it legal. The
current algorithm discards cases if the dependency is neither hoistable nor
rematerializable. We collect the \GVN{} of all the instructions in the function
and iterate on the list of instructions having identical \GVN{}s. The algorithm
\emph{doGCM} prefers hoisting (\emph{doCodeHoisting}) to sinking
(\emph{doCodeSinking}).
